---
title: "VIZ_Quarto Static DashBoard_Example1"
format: dashboard 
author: "Toluwanimi Olufawo"
---


# Page 1 

## Row 

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(readxl)  # For reading Excel files
```




```{r}
# Load the dataset
file_path <- "C:/Users/toluf/OneDrive/Desktop/week 9/Dataset_USGDP.xlsx"

data <- read_excel(file_path)

```

```{r}
# View the first few rows of the dataset
head(data)

# Check column names
colnames(data)

```
```{r}
# Load necessary library for renaming columns
library(janitor)

# Clean column names for easier reference
data <- data %>% clean_names()

# Check column names after cleaning
colnames(data)

```





# Column 1: Total U.S. GDP Over Time

```{r}
# Convert Year column to numeric if needed
data$x1 <- as.numeric(data$x1)  # Replace `x1` with the correct column name for year if it's different

# Plot Total U.S. GDP over time
ggplot(data, aes(x = x1, y = total_u_s_gdp)) +
  geom_line(color = "blue") +
  labs(title = "Total U.S. GDP Over Time",
       x = "Year",
       y = "Total U.S. GDP (in billions)") +
  theme_minimal()

```



# Column 2: For-Hire Transportation Services GDP Over Time

```{r}
# Plot For-Hire Transportation Services GDP over time
ggplot(data, aes(x = x1, y = for_hire_transportation_services_gdp_total)) +
  geom_line(color = "darkgreen") +
  labs(title = "For-Hire Transportation Services GDP Over Time",
       x = "Year",
       y = "For-Hire Transportation Services GDP (in billions)") +
  theme_minimal()

```


# Annotate Policy Impacts

```{r}

# Example policy impact data for annotations
policy_data <- data.frame(
  x1 = c(2000, 2008, 2020),  # Example years for policies
  total_u_s_gdp = c(10200, 14000, 21000),  # Replace these with approximate GDP values if known
  label = c("Policy A", "Policy B", "Policy C")
)

# Plot with annotations for policies
ggplot(data, aes(x = x1, y = total_u_s_gdp)) +
  geom_line(color = "blue") +
  geom_point(data = policy_data, aes(x = x1, y = total_u_s_gdp), color = "red", size = 3) +
  geom_text(data = policy_data, aes(x = x1, y = total_u_s_gdp, label = label), 
            vjust = -1, hjust = 0.5, color = "red") +
  labs(title = "Total U.S. GDP with Policy Impacts",
       x = "Year",
       y = "Total U.S. GDP (in billions)") +
  theme_minimal()

# Example policy impact data for annotations
policy_data <- data.frame(
  x1 = c(2000, 2008, 2020),  # Example years for policies
  total_u_s_gdp = c(10200, 14000, 21000),  # Replace these with actual or approximate GDP values if known
  label = c("Policy

```



 A", "Policy B", "Policy C")
)

# Plot with annotations for policies
gdp_policy_plot <- ggplot(data, aes(x = year, y = total_u_s_gdp)) +
  geom_line(color = "blue") +
  geom_line(aes(y = for_hire_transportation_services_gdp_total), color = "green", linetype = "dashed") + 
  geom_point(data = policy_data, aes(x = x1, y = total_u_s_gdp), color = "red", size = 3) +
  geom_text(data = policy_data, aes(x = x1, y = total_u_s_gdp, label = label), 
            vjust = -1, hjust = 0.5, color = "red") +
  labs(title = "Total U.S. GDP with Policy Impacts",
       x = "Year",
       y = "Total U.S. GDP (in billions)") +
  theme_minimal()

# Convert to interactive plot
ggplotly(gdp_policy_plot)
