---
title: "GDP Analysis Dashboard"
format: dashboard 
author: "Toluwanimi Olufawo"
---



# Page 1

## Row 1 

*Load Required Libraries*
```{r}
# Load required libraries
library(ggplot2)
library(dplyr)
library(plotly)

```

### column 
*Load and Prepare the Data*


```{r}
# Load the cleaned dataset
data <- read.csv("C:/Users/toluf/OneDrive/Desktop/week 9/Cleaned_Dataset_USGDP.csv")
```


# Page 2 

## Row 1 
Create Interactive Charts with Annotations for the sectors 
1- Air Transportation GDP Over Time

```{r}
## Air Transportation GDP Over Time

sector_plot_air <- ggplot(data, aes(x = year)) +
  geom_line(aes(y = air_transportation_4, color = "Air Transportation GDP")) +
  labs(title = "Air Transportation GDP Over Time",
       x = "Year", y = "GDP in Billions of USD") +
  theme_minimal() +
  scale_color_manual(values = c("Air Transportation GDP" = "blue"))  # Add color customization here
ggplotly(sector_plot_air)

```


## Row 2
2 - Rail Transportation GDP Over Time

```{r}
## Rail Transportation GDP Over Time

sector_plot_rail <- ggplot(data, aes(x = year)) +
  geom_line(aes(y = rail_transportation_5, color = "Rail Transportation GDP")) +
  labs(title = "Rail Transportation GDP Over Time",
       x = "Year", y = "GDP in Billions of USD") +
  theme_minimal() +
  scale_color_manual(values = c("Rail Transportation GDP" = "green"))  # Add color customization here
ggplotly(sector_plot_rail)
```



## Row 3
Water Transportation GDP Over Time
```{r}
## Water Transportation GDP Over Time

sector_plot_water <- ggplot(data, aes(x = year)) +
  geom_line(aes(y = water_transportation_6, color = "Water Transportation GDP")) +
  labs(title = "Water Transportation GDP Over Time",
       x = "Year", y = "GDP in Billions of USD") +
  theme_minimal() +
  scale_color_manual(values = c("Water Transportation GDP" = "pink"))  # Optional color customization
ggplotly(sector_plot_water)
```
## Row 4
Truck Transportation GDP Over Time
```{r}
## Truck Transportation GDP Over Time

sector_plot_truck <- ggplot(data, aes(x = year)) +
  geom_line(aes(y = truck_transportation_7, color = "Truck Transportation GDP")) +
  labs(title = "Truck Transportation GDP Over Time",
       x = "Year", y = "GDP in Billions of USD") +
  theme_minimal() +
  scale_color_manual(values = c("Truck Transportation GDP" = "orange"))  # Optional color customization
ggplotly(sector_plot_truck)

```


# Page 3

## Row 1
list of sectors

```{r}
# Define a list of sectors and their respective column names
sectors <- list(
  "Transit and Ground Passenger GDP" = "transit_and_ground_passenger_transportation_8",
  "Pipeline Transportation GDP" = "pipeline_transportation_9",
  "Other Transportation Support GDP" = "other_transportation_and_support_activities_10",
  "Warehousing and Storage GDP" = "warehousing_and_storage_11"
)

# Generate a plot for each sector
for (sector in names(sectors)) {
  column_name <- sectors[[sector]]
  
  plot <- ggplot(data, aes(x = year)) +
    geom_line(aes_string(y = column_name, color = shQuote(sector))) +
    labs(title = paste(sector, "Over Time"),
         x = "Year", y = "GDP in Billions of USD") +
    theme_minimal()
  
  print(ggplotly(plot))
}
```


## Row 2- Annotated GDP Growth Chart

```{r}
## GDP Growth Over Time with Policy Annotations


# Example policy impact data
policy_impacts <- data.frame(
  Year = c(1990, 2001, 2008),  # Replace these with the actual policy years
  Policy = c("Policy A", "Policy B", "Policy C")
)

# Create the plot
gdp_plot <- ggplot(data, aes(x = year)) +
  geom_line(aes(y = total_u_s_gdp, color = "Total U.S. GDP")) +
  geom_line(aes(y = for_hire_transportation_services_gdp_total, color = "For-Hire Transportation GDP")) +
  geom_text(data = policy_impacts, aes(x = Year, y = max(data$total_u_s_gdp), label = Policy),
            vjust = -1, color = "black", angle = 45) +
  labs(title = "GDP Contribution Over Time with Policy Annotations",
       x = "Year", y = "GDP in Billions of USD") +
  theme_minimal()

# Convert to an interactive plot
ggplotly(gdp_plot)


`````



# Page  4.
## Row 1
Additional Sections for Different Sectors

To visualize GDP contributions by different transportation sectors

```{r}
## GDP by Transportation Sector
# Example for plotting Air Transportation GDP over time
sector_plot <- ggplot(data, aes(x = year)) +
  geom_line(aes(y = air_transportation_4, color = "Air Transportation GDP")) +
  labs(title = "Air Transportation GDP Over Time",
       x = "Year", y = "GDP in Billions of USD") +
  theme_minimal()

# Convert to interactive plot
ggplotly(sector_plot)
`````



# Page 5 
## Create the Chart Section

Add a new section to visualize **Total U.S. GDP with Policy Impacts**.

```{r}
# Total U.S. GDP with Policy Impacts
# Example of policy impact data for annotations
policy_impacts <- data.frame(
  year = c(1990, 2001, 2008),  # Replace with actual years of policy impacts
  Policy = c("Policy A", "Policy B", "Policy C")
)

 #Create the U.S. GDP plot with annotations for policy impacts
gdp_plot <- ggplot(data, aes(x = year, y = total_u_s_gdp)) +
  geom_line(color = "blue", size = 1) +
  geom_point(data = policy_impacts, aes(x = year, y = max(data$total_u_s_gdp)), 
             color = "red", size = 3) +
  geom_text(data = policy_impacts, aes(x = year, y = max(data$total_u_s_gdp), label = Policy),
            vjust = -1, color = "black", angle = 45, size = 3) +
  labs(title = "Total U.S. GDP with Policy Impacts",
       x = "Year",
       y = "GDP in Billions of USD") +
  theme_minimal()

# Convert the plot to an interactive plot
ggplotly(gdp_plot)

````









