# 1. Abigail Tako
### Week 1 
# 2. Caleb Pena
### Week 1 
# 3. Derrick Baruga
### Week 1 
# 4. Toluwanimi Olufawo 
### Week 1

**To-Do list for this week**
Develop a timeline outlining major external factors (fuel prices, policy changes) affecting the transportation sector and visualize it in R

**Load Required Libraries**
```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(readxl)  # For reading Excel files

```

**Load the Dataset**
```{r}
# Load the dataset
file_path <- "C:/Users/toluf/OneDrive/Desktop/Dynamic 2/Olufawo_FinalProjectWeek1/Dataset_USGDP.xlsx"
data <- read_excel(file_path)

```
**Inspect and Clean the Data**
```{r}
# View the first few rows of the dataset
head(data)

# Check column names
colnames(data)

```



**Filter Data for Timeline Events**
Since I'm focused on creating a timeline of external factors like fuel prices and policy changes, I'll need to isolate specific years and events to include in the timeline. I’ll add a subset of relevant data and then create a separate data frame for these events.
```{r}
# Create a data frame with specific events for the timeline
events <- data.frame(
  year = c(2005, 2008, 2009, 2010, 2011, 2012, 2015, 2020, 2021, 2022),
  label = c(
    "Fuel Prices Spike",
    "Energy Policy Act",
    "Global Financial Crisis",
    "FMCSA Safety Regulations",
    "Natural Gas Boom",
    "MAP-21 Act",
    "Low Oil Prices",
    "COVID-19 Pandemic & CARES Act",
    "Supply Chain Disruptions & Infrastructure Act",
    "Fuel Price Surge"
  ),
  type = c("Fuel", "Policy", "Global", "Policy", "Fuel", "Policy", "Fuel", "Global", "Global/Policy", "Fuel")
)

```

**Plot the Timeline**
```{r}
# Plot the timeline with line width aesthetic
ggplot(events, aes(x = year, y = 1, color = type)) +
  geom_point(size = 5) +  # Points for each event
  geom_segment(aes(x = year, xend = year, y = 1, yend = 0.9), color = "gray", linewidth = 0.5) +  # Adjusted to use linewidth
  geom_text(aes(label = label), angle = 45, hjust = 0, nudge_y = 0.2, size = 3.5) +  # Event labels
  scale_x_continuous(breaks = seq(2005, 2022, by = 1)) +  # Set x-axis breaks for each year
  labs(
    title = "Timeline of Major External Factors Affecting the U.S. Transportation Sector",
    x = "Year",
    y = ""
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),  # Remove y-axis text
    axis.ticks.y = element_blank(),  # Remove y-axis ticks
    legend.position = "bottom"
  ) +
  guides(color = guide_legend(title = "Event Type"))  # Legend for event types


```



**Modify the Data for a Gantt-Style Chart**



```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Define the data with start and end years for Gantt-style visualization
events <- data.frame(
  event = c(
    "Fuel Prices Spike", "Energy Policy Act", "Global Financial Crisis",
    "FMCSA Safety Regulations", "Natural Gas Boom", "MAP-21 Act",
    "Low Oil Prices", "COVID-19 Pandemic & CARES Act", 
    "Supply Chain Disruptions & Infrastructure Act", "Fuel Price Surge"
  ),
  type = c("Fuel", "Policy", "Global", "Policy", "Fuel", "Policy", "Fuel", "Global", "Global/Policy", "Fuel"),
  start_year = c(2005, 2008, 2009, 2010, 2011, 2012, 2015, 2020, 2021, 2022),
  end_year = c(2008, 2008, 2009, 2010, 2014, 2012, 2018, 2020, 2022, 2022)
)

```

**Create the Gantt-Style Timeline Chart**


```{r}
# Gantt-style timeline plot
ggplot(events, aes(x = start_year, xend = end_year, y = event, yend = event, color = type)) +
  geom_segment(size = 5) +  # Create horizontal bars for each event
  scale_x_continuous(breaks = seq(2005, 2022, by = 1)) +  # Set x-axis breaks for each year
  labs(
    title = "Gantt-Style Timeline of Major External Factors Affecting the U.S. Transportation Sector",
    x = "Year",
    y = "Event"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8),  # Adjust y-axis text size for readability
    legend.position = "bottom"
  ) +
  guides(color = guide_legend(title = "Event Type"))  # Legend for event types

```
