---
title: "Untitled"
format: html
editor: visual
---

# 1.Load Data and Packages

```{r}

# Load libraries
library(ggplot2)
library(dplyr)
library(tidyr)

# Load the dataset
data <- read.csv("C:/Users/toluf/OneDrive/Desktop/Week_10/Cleaned_Dataset_USGDP.csv")


```

# 2. Plot Total GDP and Transportation GDP Timeline

```{r}
# Plot Total GDP vs Transportation GDP over Time
ggplot(data, aes(x = year)) +
  geom_line(aes(y = total_u_s_gdp, color = "Total U.S. GDP")) +
  geom_line(aes(y = for_hire_transportation_services_gdp_total, color = "For-Hire Transportation Services GDP")) +
  labs(title = "Total U.S. GDP vs For-Hire Transportation Services GDP Over Time",
       x = "Year", y = "GDP (in billions of current dollars)") +
  scale_color_manual("", values = c("Total U.S. GDP" = "blue", "For-Hire Transportation Services GDP" = "red")) +
  theme_minimal()

```

```         
Observation: The blue line representing total U.S. GDP shows a steady and significant increase, especially after the 2000s, while the red line for for-hire transportation services GDP grows at a much slower rate

Interpretation:

Historical Context and Policies:

- Deregulation: In the 1980s, deregulation policies like the Motor Carrier Act (1980) and the Staggers Rail Act (1980) reduced government control over the rail and trucking industries, allowing these sectors to become more cost-effective. However, deregulation also led to increased competition, which may explain the slower growth rate of transportation GDP compared to the broader economy.

- Technological Advances: Technological improvements, especially in logistics management, helped maintain stable growth in transportation GDP. Automation, electronic data interchange (EDI), and more efficient scheduling contributed to keeping costs down but also limited transportation sector expansion in relation to the total GDP.

- Globalization and Trade: Growth in trade agreements (such as NAFTA in 1994) increased demand for freight transport but benefited other sectors of the economy disproportionately, resulting in a moderate contribution of transportation services to total GDP.

Explanation of Trends:
- While the overall U.S. GDP grew rapidly due to diversification and growth in technology, finance, and healthcare, the transportation sector grew at a slower rate because its expansion is more closely tied to physical goods movement, which has limits in terms of growth relative to GDP-driven sectors like technology..
```

3.  Plot Sector-Specific Trends Over Time

```{r}
# Gather data for individual sectors to plot them separately
data_long <- data %>%
  select(year, air_transportation_4, rail_transportation_5, water_transportation_6, truck_transportation_7) %>%
  gather(key = "sector", value = "gdp", -year)

# Plot sector-specific trends over time
ggplot(data_long, aes(x = year, y = gdp, color = sector)) +
  geom_line() +
  labs(title = "Sector-Specific GDP Trends in For-Hire Transportation Services",
       x = "Year", y = "GDP (in billions of current dollars)") +
  scale_color_discrete(name = "Sector", labels = c("Air", "Rail", "Water", "Truck")) +
  theme_minimal()

```

```         
Observation: Each line represents a different transportation sector’s GDP over time. The lines for sectors like truck and air transportation show the most significant growth, while others, such as rail and water transportation, have relatively flat trends.

Interpretation:

Historical Context and Policies:

- Truck Transportation: Deregulation and the rise of e-commerce have continuously boosted the demand for trucking. The increased reliance on trucks for last-mile deliveries has reinforced trucking’s dominance within the sector. Federal Highway Administration (FHWA) policies have also invested heavily in highway infrastructure supporting trucking expansion.

- Air Transportation: Rising global trade and policies supporting air cargo infrastructure (such as improved logistics at major airports) contributed to the increase in air transportation GDP. Since the 2000s, advancements in air freight services for time-sensitive and high-value goods (e.g., electronics) have been significant.

- Rail and Water Transport: Despite being efficient for bulk goods, rail and water have shown slower growth due to competition from trucking and air freight. Rail freight has particularly struggled with intermodal competition, and regulatory policies that focus on maintaining safety standards without sufficient investments in infrastructure modernization have limited growth.

Explanation of Trends:

- The steep rise in trucking reflects increased demand for flexible and responsive shipping solutions, especially for the e-commerce boom. Air transportation's growth is driven by the need for speed and high-value shipments, while the minimal growth in rail and water reflects their specialized use and limited flexibility compared to trucking.
```

4.Plot Percentage Contribution of Each Sector

```{r}
# Select only relevant percentage columns for each sector
percentage_data <- data %>%
  select(year, air_transportation_23, rail_transportation_24, water_transportation_25,
         truck_transportation_26, transit_and_ground_passenger_transportation_27,
         pipeline_transportation_28, other_transportation_and_support_activities_29,
         warehousing_and_storage_30) %>%
  gather(key = "sector", value = "percentage", -year)

# Plot percentage contribution over time
ggplot(percentage_data, aes(x = year, y = percentage, color = sector)) +
  geom_line() +
  labs(title = "Percentage Contribution of Each Sector in For-Hire Transportation Services GDP",
       x = "Year", y = "Percentage of Total For-Hire Transportation GDP") +
  theme_minimal() +
  scale_color_discrete(name = "Sector")

```

```         
Historical Context and Policies:

- Truck Transportation Dominance: Deregulation, infrastructure investment, and the rise of logistics-driven business models have made trucking essential for goods distribution across the U.S. In addition, policies have favored the trucking industry with investments in road infrastructure, supporting its role as a primary transportation mode.

- Air Transportation Growth: Policies promoting airport modernization and international trade agreements have allowed air cargo to grow. The increasing share of air transport is aligned with global trade growth and demand for rapid transport of high-value goods.

- Decline in Rail and Water Contribution: Rail, once essential for bulk goods transport, has faced operational and investment challenges. The limited expansion of rail infrastructure and competition with trucking for shorter, regional hauls have reduced its share in GDP. Water transportation has been similarly limited, with policies more focused on environmental regulations than expansion, limiting its growth in the transportation GDP.

Explanation of Trends:

The stable and high contribution of truck transportation reflects its adaptability and infrastructure support, while air transportation’s growth aligns with the demand for faster logistics. Rail and water, facing operational constraints and lack of significant infrastructure investment, have become less dominant contributors to transportation GDP.
```

5.Draft Impact Analysis Insights

```{r}
# Calculate annual growth rates for total and sectoral GDPs
data <- data %>%
  mutate(total_gdp_growth = (total_u_s_gdp - lag(total_u_s_gdp)) / lag(total_u_s_gdp) * 100,
         transport_gdp_growth = (for_hire_transportation_services_gdp_total - lag(for_hire_transportation_services_gdp_total)) / lag(for_hire_transportation_services_gdp_total) * 100)

# Print growth rate data for analysis
print(data %>% select(year, total_gdp_growth, transport_gdp_growth))

```

```         
Summary of Policy Impact on Each Mode of Transport

1.Truck Transportation:

- Impact of Deregulation: Deregulation in the 1980s allowed for a competitive, cost-efficient trucking industry, leading to its substantial share in transportation GDP.

- Infrastructure Investment: Consistent investment in the U.S. interstate highway system has reinforced the trucking sector’s dominance, enabling it to meet the demands of e-commerce and last-mile logistics.

- Current Trends: Trucking continues to benefit from policy emphasis on infrastructure modernization and congestion management, making it the most resilient mode in response to economic shifts.

2. Air Transportation:

- Globalization and High-Value Goods: As trade expanded, air freight became more significant for time-sensitive, high-value items. Policies promoting airport infrastructure modernization have further supported this growth.

- Future Potential: Continued investment in air freight infrastructure and technology could increase air transportation’s GDP contribution, particularly as demand grows for international trade.

3. Rail and Water Transportation:

- Specialization and Regulatory Challenges: While rail is efficient for bulk goods, regulatory emphasis on safety over infrastructure modernization has limited its flexibility and growth. Water transport, similarly constrained by environmental policies, plays a role but remains less dynamic.

- Competition with Trucking: Rail’s decline relative to trucking reflects competitive disadvantages in flexibility and speed, with rail primarily limited to specialized cargo. Similarly, water transport is constrained by geographical and regulatory limitations.
 
These insights provide a comprehensive understanding of how policies and economic trends have shaped each mode’s contribution to transportation GDP over time. These details can be used to add historical and policy context to each graph in your presentation. Let me know if you need specific sources or more details on any particular point
```

# Timeline outlining major external factors

```{r}
# Load libraries
library(plotly)
library(dplyr)
```

## 1.Define the Data for the Timeline

```{r}
# Define data for the timeline to create a dataframe with key dates, events, descriptions, and categories.
timeline_data <- data.frame(
  Year = c(1973, 1978, 1980, 1980, 1985, 1994, 2001, 2008, 2010, 2015, 2020, 2021),
  Event = c(
    "1973 Oil Crisis",
    "Airline Deregulation Act",
    "Motor Carrier Act",
    "Staggers Rail Act",
    "Decline in Fuel Prices",
    "NAFTA Agreement",
    "Rise in Fuel Prices",
    "2008 Recession",
    "Clean Truck Program",
    "Diesel Price Decline",
    "COVID-19 Pandemic",
    "Infrastructure Investment and Jobs Act"
  ),
  Description = c(
    "OPEC oil embargo led to sharp increase in fuel prices.",
    "Deregulated the airline industry, leading to increased competition.",
    "Deregulated trucking industry, allowing for increased competition.",
    "Deregulated rail industry, allowing rail companies to set competitive rates.",
    "Fuel prices began to decline, reducing operational costs.",
    "Increased cross-border trucking and rail traffic between U.S., Canada, and Mexico.",
    "Steady increase in fuel prices during this period, peaking in 2008.",
    "Global economic downturn reduced demand across the transportation sector.",
    "Stricter environmental standards for trucks, encouraging cleaner technologies.",
    "Diesel prices fell due to increased U.S. oil production.",
    "Pandemic disrupted global supply chains and shifted demand.",
    "Allocated significant funding for transportation infrastructure improvements."
  ),
  Category = c(
    "Fuel Crisis", "Policy Change", "Policy Change", "Policy Change", "Fuel Prices", 
    "Policy Change", "Fuel Prices", "Economic Event", "Environmental Policy", 
    "Fuel Prices", "Pandemic", "Infrastructure Investment"
  )
)
# Assign colors to each category for easy distinction
colors <- c("Fuel Crisis" = "red", "Policy Change" = "blue", "Fuel Prices" = "orange",
            "Economic Event" = "green", "Environmental Policy" = "purple", 
            "Pandemic" = "pink", "Infrastructure Investment" = "brown")

# Create interactive plot with color-coded categories
timeline_plot <- plot_ly(
  data = timeline_data,
  x = ~Year,
  y = ~Event,
  type = "scatter",
  mode = "markers+text",
  text = ~paste("<b>Event:</b>", Event, "<br><b>Description:</b>", Description, "<br><b>Category:</b>", Category),
  hoverinfo = "text",
  marker = list(size = 10),
  color = ~Category,
  colors = colors
) %>%
  layout(
    title = "Timeline of Major External Factors Affecting the U.S. Transportation Sector",
    xaxis = list(title = "Year", tickmode = "linear"),
    yaxis = list(title = "", autorange = "reversed"),
    hoverlabel = list(bgcolor = "white", font = list(size = 12)),
    showlegend = TRUE
  )

# Display the plot
timeline_plot
```

## 2. Create an Interactive Plotly Timeline

### 2.1 Assign Nicknames for every major event

Year Event Nickname 1973 1973 Oil Crisis OIL-73 1978 Airline Deregulation Act AIR-DEREG 1980 Motor Carrier Act TRK-DEREG 1980 Staggers Rail Act RAIL-DEREG 1985 Decline in Fuel Prices FUEL-DECL 1994 NAFTA Agreement NAFTA 2001 Rise in Fuel Prices FUEL-RISE 2008 2008 Recession REC-08 2010 Clean Truck Program TRK-CLEAN 2015 Diesel Price Decline DIESEL-DECL 2020 COVID-19 Pandemic COVID-19 2021 Infrastructure Investment and Jobs Act INFRA-21

### 2.2 Prepare Data by Decades

```{r}

# Updated data with nicknames
timeline_data <- data.frame(
  Year = c(1973, 1978, 1980, 1980, 1985, 1994, 2001, 2008, 2010, 2015, 2020, 2021),
  Nickname = c("OIL-73", "AIR-DEREG", "TRK-DEREG", "RAIL-DEREG", "FUEL-DECL",
               "NAFTA", "FUEL-RISE", "REC-08", "TRK-CLEAN", "DIESEL-DECL", 
               "COVID-19", "INFRA-21"),
  Description = c(
    "OPEC oil embargo led to sharp increase in fuel prices.",
    "Deregulated the airline industry, leading to increased competition.",
    "Deregulated trucking industry, allowing for increased competition.",
    "Deregulated rail industry, allowing rail companies to set competitive rates.",
    "Fuel prices began to decline, reducing operational costs.",
    "Increased cross-border trucking and rail traffic between U.S., Canada, and Mexico.",
    "Steady increase in fuel prices during this period, peaking in 2008.",
    "Global economic downturn reduced demand across the transportation sector.",
    "Stricter environmental standards for trucks, encouraging cleaner technologies.",
    "Diesel prices fell due to increased U.S. oil production.",
    "Pandemic disrupted global supply chains and shifted demand.",
    "Allocated significant funding for transportation infrastructure improvements."
  ),
  Category = c(
    "Fuel Crisis", "Policy Change", "Policy Change", "Policy Change", "Fuel Prices", 
    "Policy Change", "Fuel Prices", "Economic Event", "Environmental Policy", 
    "Fuel Prices", "Pandemic", "Infrastructure Investment"
  )
)

# Split data by decades
timeline_1970s <- subset(timeline_data, Year < 1980)
timeline_1980s <- subset(timeline_data, Year >= 1980 & Year < 1990)
timeline_1990s <- subset(timeline_data, Year >= 1990 & Year < 2000)
timeline_2000s <- subset(timeline_data, Year >= 2000 & Year < 2010)
timeline_2010s <- subset(timeline_data, Year >= 2010)


```

##### Trials

Timeline with Annotations

```{r}


library(ggplot2)
library(ggrepel)

# Plot with annotations
ggplot(timeline_data, aes(x = Year, y = 0, color = Category)) +
  geom_point(size = 4) +
  geom_text_repel(aes(label = Nickname), size = 3, direction = "y", nudge_y = 0.2) +
  scale_color_manual(values = colors) +
  labs(title = "Timeline of Major External Factors Affecting the U.S. Transportation Sector",
       x = "Year", y = "") +
  theme_minimal() +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.minor.y = element_blank(), legend.position = "bottom")

```

```{r}
library(plotly)
library(dplyr)

# Define data with categories, nicknames, and colors
timeline_data <- data.frame(
  Year = c(1973, 1978, 1980, 1980, 1985, 1994, 2001, 2008, 2010, 2015, 2020, 2021),
  Nickname = c("OIL-73", "AIR-DEREG", "TRK-DEREG", "RAIL-DEREG", "FUEL-DECL",
               "NAFTA", "FUEL-RISE", "REC-08", "TRK-CLEAN", "DIESEL-DECL", 
               "COVID-19", "INFRA-21"),
  Description = c(
    "OPEC oil embargo led to sharp increase in fuel prices.",
    "Deregulated the airline industry, leading to increased competition.",
    "Deregulated trucking industry, allowing for increased competition.",
    "Deregulated rail industry, allowing rail companies to set competitive rates.",
    "Fuel prices began to decline, reducing operational costs.",
    "Increased cross-border trucking and rail traffic between U.S., Canada, and Mexico.",
    "Steady increase in fuel prices during this period, peaking in 2008.",
    "Global economic downturn reduced demand across the transportation sector.",
    "Stricter environmental standards for trucks, encouraging cleaner technologies.",
    "Diesel prices fell due to increased U.S. oil production.",
    "Pandemic disrupted global supply chains and shifted demand.",
    "Allocated significant funding for transportation infrastructure improvements."
  ),
  Category = c(
    "Fuel Crisis", "Policy Change", "Policy Change", "Policy Change", "Fuel Prices", 
    "Policy Change", "Fuel Prices", "Economic Event", "Environmental Policy", 
    "Fuel Prices", "Pandemic", "Infrastructure Investment"
  )
)

# Assign colors to each category for easy distinction
colors <- c("Fuel Crisis" = "red", "Policy Change" = "blue", "Fuel Prices" = "orange",
            "Economic Event" = "green", "Environmental Policy" = "purple", 
            "Pandemic" = "pink", "Infrastructure Investment" = "brown")

# Interactive plot with annotations
timeline_plot <- plot_ly(
  data = timeline_data,
  x = ~Year,
  y = ~0,  # keep all points on the same y-axis level for a horizontal timeline
  type = "scatter",
  mode = "markers+text",
  text = ~paste("<b>Event:</b>", Nickname, "<br><b>Description:</b>", Description, "<br><b>Category:</b>", Category),
  hoverinfo = "text",
  marker = list(size = 10),
  color = ~Category,
  colors = colors,
  textposition = "top center",
  textfont = list(size = 10, color = "black")
) %>%
  layout(
    title = "Interactive Timeline of Major External Factors Affecting the U.S. Transportation Sector",
    xaxis = list(title = "Year", tickmode = "linear"),
    yaxis = list(title = "", showticklabels = FALSE),  # hide y-axis labels
    hoverlabel = list(bgcolor = "white", font = list(size = 12)),
    showlegend = TRUE,
    legend = list(orientation = "h", x = 0.5, xanchor = "center", y = -0.2)  # position legend below the plot
  )

# Display the plot
timeline_plot

```

```{r}
library(plotly)
library(dplyr)

# Define data with categories, nicknames, and colors
timeline_data <- data.frame(
  Year = c(1973, 1978, 1980, 1980, 1985, 1994, 2001, 2008, 2010, 2015, 2020, 2021),
  Nickname = c("OIL-73", "AIR-DEREG", "TRK-DEREG", "RAIL-DEREG", "FUEL-DECL",
               "NAFTA", "FUEL-RISE", "REC-08", "TRK-CLEAN", "DIESEL-DECL", 
               "COVID-19", "INFRA-21"),
  Description = c(
    "OPEC oil embargo led to sharp increase in fuel prices.",
    "Deregulated the airline industry, leading to increased competition.",
    "Deregulated trucking industry, allowing for increased competition.",
    "Deregulated rail industry, allowing rail companies to set competitive rates.",
    "Fuel prices began to decline, reducing operational costs.",
    "Increased cross-border trucking and rail traffic between U.S., Canada, and Mexico.",
    "Steady increase in fuel prices during this period, peaking in 2008.",
    "Global economic downturn reduced demand across the transportation sector.",
    "Stricter environmental standards for trucks, encouraging cleaner technologies.",
    "Diesel prices fell due to increased U.S. oil production.",
    "Pandemic disrupted global supply chains and shifted demand.",
    "Allocated significant funding for transportation infrastructure improvements."
  ),
  Category = c(
    "Fuel Crisis", "Policy Change", "Policy Change", "Policy Change", "Fuel Prices", 
    "Policy Change", "Fuel Prices", "Economic Event", "Environmental Policy", 
    "Fuel Prices", "Pandemic", "Infrastructure Investment"
  )
)

# Assign colors to each category for easy distinction
colors <- c("Fuel Crisis" = "red", "Policy Change" = "blue", "Fuel Prices" = "orange",
            "Economic Event" = "green", "Environmental Policy" = "purple", 
            "Pandemic" = "pink", "Infrastructure Investment" = "brown")

# Interactive plot without text labels, showing details only on hover
timeline_plot <- plot_ly(
  data = timeline_data,
  x = ~Year,
  y = ~0,  # keep all points on the same y-axis level for a horizontal timeline
  type = "scatter",
  mode = "markers",
  marker = list(size = 12, opacity = 0.8, line = list(width = 2, color = 'black')),
  color = ~Category,
  colors = colors,
  hoverinfo = "text",
  text = ~paste("<b>Event:</b>", Nickname, "<br><b>Description:</b>", Description, "<br><b>Category:</b>", Category)
) %>%
  layout(
    title = list(
      text = "<b>Interactive Timeline of Major External Factors Affecting the U.S. Transportation Sector</b>",
      font = list(size = 18)
    ),
    xaxis = list(
      title = "Year",
      tickmode = "linear",
      showgrid = FALSE,
      zeroline = FALSE
    ),
    yaxis = list(
      title = "",
      showticklabels = FALSE,
      zeroline = FALSE
    ),  # hide y-axis labels
    hoverlabel = list(
      bgcolor = "white",
      font = list(size = 12)
    ),
    showlegend = TRUE,
    legend = list(
      orientation = "h",
      x = 0.5, xanchor = "center",
      y = -0.2
    )  # position legend below the plot
  )

# Display the plot
timeline_plot

```

```{r}
library(plotly)
library(dplyr)

# Load the GDP dataset
gdp_data <- read.csv("C:/Users/toluf/OneDrive/Desktop/Week_10/Cleaned_Dataset_USGDP.csv")

# Prepare the timeline data with added details for each year
timeline_data <- data.frame(
  Year = gdp_data$year,
  Total_GDP = gdp_data$total_u_s_gdp,
  Transportation_GDP = gdp_data$for_hire_transportation_services_gdp_total,
  Air = gdp_data$air_transportation_4,
  Rail = gdp_data$rail_transportation_5,
  Water = gdp_data$water_transportation_6,
  Truck = gdp_data$truck_transportation_7,
  Transit = gdp_data$transit_and_ground_passenger_transportation_8,
  Pipeline = gdp_data$pipeline_transportation_9,
  Support = gdp_data$other_transportation_and_support_activities_10,
  Warehousing = gdp_data$warehousing_and_storage_21,
  Percent_Transport_GDP = gdp_data$percent_of_for_hire_transportation_services_gdp
)

# Create the interactive plot with enhanced hover information and refined layout
timeline_plot <- plot_ly(
  data = timeline_data,
  x = ~Year,
  y = ~0,  # keep all points on the same y-axis level for a horizontal timeline
  type = "scatter",
  mode = "markers",
  marker = list(size = 12, opacity = 0.8, line = list(width = 2, color = 'black')),
  color = ~Year,  # Optional: Use a color scale by year to visually distinguish them
  hoverinfo = "text",
  text = ~paste(
    "<b>Year:</b>", Year,
    "<br><b>Total U.S. GDP:</b> $", round(Total_GDP, 1), " billion",
    "<br><b>Transportation GDP:</b> $", round(Transportation_GDP, 1), " billion",
    "<br><b>Air:</b> $", round(Air, 1), " billion",
    "<br><b>Rail:</b> $", round(Rail, 1), " billion",
    "<br><b>Water:</b> $", round(Water, 1), " billion",
    "<br><b>Truck:</b> $", round(Truck, 1), " billion",
    "<br><b>Transit:</b> $", round(Transit, 1), " billion",
    "<br><b>Pipeline:</b> $", round(Pipeline, 1), " billion",
    "<br><b>Support Activities:</b> $", round(Support, 1), " billion",
    "<br><b>Warehousing:</b> $", round(Warehousing, 1), " billion",
    "<br><b>Transportation as % of Total GDP:</b> ", round(Percent_Transport_GDP, 1), "%"
  )
) %>%
  layout(
    title = list(
      text = "<b>Interactive Timeline of GDP Contributions by Transportation Sector</b>",
      font = list(size = 18)
    ),
    xaxis = list(
      title = "Year",
      tickmode = "linear",
      showgrid = FALSE,
      zeroline = FALSE,
      tickangle = 45,  # Rotate x-axis labels for better readability
      dtick = 5  # Display every 5 years to reduce label clutter
    ),
    yaxis = list(
      title = "",
      showticklabels = FALSE,
      zeroline = FALSE
    ),  # hide y-axis labels
    hoverlabel = list(
      bgcolor = "white",
      font = list(size = 12)
    ),
    showlegend = FALSE
  )

# Display the plot
timeline_plot
```

Timeline of Major External Factors Affecting the U.S. Transportation Sector

1970s • 1973-1974 Oil Crisis: The OPEC oil embargo led to a sharp increase in fuel prices, impacting transportation costs across all modes. This crisis forced companies to adopt fuel-saving practices and consider alternative transportation modes to reduce dependence on fuel. • 1978 Airline Deregulation Act: This act deregulated the airline industry, leading to increased competition, lower airfares, and a surge in air freight activity as airlines explored new business opportunities. 1980s • 1980 Motor Carrier Act: Deregulated the trucking industry, allowing for increased competition, lower rates, and more efficient service. This policy was instrumental in the rise of trucking as the primary mode for goods distribution within the U.S. • 1980 Staggers Rail Act: Deregulated the rail industry, allowing rail companies to set competitive rates and abandon unprofitable routes, which revitalized the rail freight sector and enabled it to become more cost-effective for bulk goods transport. • 1980-1985 Decline in Fuel Prices: Following the high fuel prices of the 1970s, fuel prices began to decline in the 1980s, reducing operational costs across all transportation modes, particularly benefiting truck and air transportation. 1990s • NAFTA (1994): The North American Free Trade Agreement increased trade between the U.S., Canada, and Mexico, leading to a surge in cross-border trucking and rail traffic. This agreement made North America more integrated economically, benefiting freight transportation. • Technological Advancements in Logistics: The 1990s saw the rise of electronic data interchange (EDI) and early logistics management software, which improved efficiency and coordination in the supply chain, particularly for truck transportation. • Stable Fuel Prices: For most of the 1990s, fuel prices were relatively stable, which allowed transportation companies to focus on expansion and operational improvements rather than adjusting for fuel cost volatility. 2000s • 2001-2008 Rise in Fuel Prices: A steady increase in fuel prices during this period, culminating in a peak around 2008, led to higher operating costs for all transportation sectors. This forced many companies to adopt fuel surcharges and explore fuel-efficient technologies. • Growth of E-commerce: With the rise of companies like Amazon, demand for last-mile and regional transportation increased, heavily boosting the trucking and air freight industries to meet fast shipping expectations. • 2008 Recession: The global economic downturn reduced demand across the transportation sector, particularly affecting freight volumes. However, this period also accelerated consolidation and cost-cutting strategies within the sector. 2010s • 2010 Clean Truck Program (California): Aimed to reduce emissions from trucks by enforcing stricter environmental standards. Similar programs in other states increased costs for trucking companies but also led to investments in cleaner technologies. • 2015 Diesel Price Decline: Following years of high fuel prices, diesel prices fell significantly after 2014 due to increased U.S. oil production. This decline in fuel costs lowered operating expenses for the trucking sector, making it more competitive. • 2015 Fixing America’s Surface Transportation (FAST) Act: Provided long-term funding for highway infrastructure improvements, supporting the trucking industry by enhancing road quality and reducing congestion. • Autonomous and Electric Vehicles: The 2010s saw rapid advancements in autonomous driving technology and electric vehicle development, particularly affecting trucking and warehousing. While the technology was not widely adopted yet, it set the stage for future changes in the transportation industry. 2020s • COVID-19 Pandemic (2020): The pandemic disrupted global supply chains, leading to unprecedented challenges and demand shifts within the transportation sector. Air and truck transport saw spikes in demand due to the need for rapid delivery of essential goods, while passenger transportation plummeted. • 2021 Fuel Price Surge: As economies reopened after COVID-19 lockdowns, fuel prices surged due to supply chain disruptions, impacting operating costs for all transportation sectors. • Infrastructure Investment and Jobs Act (2021): This act allocated significant funding for transportation infrastructure, including road, rail, and port improvements. It is expected to benefit the trucking industry by reducing congestion and improving road quality, while also supporting rail and waterway infrastructure. • Rise of E-commerce and Direct-to-Consumer Models: The continued growth of e-commerce has increased the demand for last-mile delivery solutions, benefiting the trucking and warehousing sectors. Consumers' preference for faster, direct deliveries has shifted supply chain strategies toward more localized warehousing and distribution.

```{r}
library(plotly)

# Prepare the GDP timeline data with added details for each year
timeline_data <- data.frame(
  Year = gdp_data$year,
  Total_GDP = gdp_data$total_u_s_gdp,
  Transportation_GDP = gdp_data$for_hire_transportation_services_gdp_total,
  Air = gdp_data$air_transportation_4,
  Rail = gdp_data$rail_transportation_5,
  Water = gdp_data$water_transportation_6,
  Truck = gdp_data$truck_transportation_7,
  Transit = gdp_data$transit_and_ground_passenger_transportation_8,
  Pipeline = gdp_data$pipeline_transportation_9,
  Support = gdp_data$other_transportation_and_support_activities_10,
  Warehousing = gdp_data$warehousing_and_storage_21,
  Percent_Transport_GDP = gdp_data$percent_of_for_hire_transportation_services_gdp
)

# Define the external factors data
external_factors <- data.frame(
  Year = c(1973, 1978, 1980, 1980, 1985, 1994, 2001, 2008, 2010, 2015, 2020, 2021),
  Event = c("OIL-73", "AIR-DEREG", "TRK-DEREG", "RAIL-DEREG", "FUEL-DECL",
            "NAFTA", "FUEL-RISE", "REC-08", "TRK-CLEAN", "DIESEL-DECL", 
            "COVID-19", "INFRA-21"),
  Description = c(
    "OPEC oil embargo led to sharp increase in fuel prices.",
    "Deregulated the airline industry, leading to increased competition.",
    "Deregulated trucking industry, allowing for increased competition.",
    "Deregulated rail industry, allowing rail companies to set competitive rates.",
    "Fuel prices began to decline, reducing operational costs.",
    "Increased cross-border trucking and rail traffic between U.S., Canada, and Mexico.",
    "Steady increase in fuel prices during this period, peaking in 2008.",
    "Global economic downturn reduced demand across the transportation sector.",
    "Stricter environmental standards for trucks, encouraging cleaner technologies.",
    "Diesel prices fell due to increased U.S. oil production.",
    "Pandemic disrupted global supply chains and shifted demand.",
    "Allocated significant funding for transportation infrastructure improvements."
  ),
  Category = c(
    "Fuel Crisis", "Policy Change", "Policy Change", "Policy Change", "Fuel Prices", 
    "Policy Change", "Fuel Prices", "Economic Event", "Environmental Policy", 
    "Fuel Prices", "Pandemic", "Infrastructure Investment"
  )
)

# Create the interactive GDP timeline with external events overlay
timeline_plot <- plot_ly() %>%
  # GDP data layer
  add_trace(
    data = timeline_data,
    x = ~Year,
    y = ~0,  # keep all GDP points on the same y-axis level
    type = "scatter",
    mode = "markers",
    marker = list(size = 12, opacity = 0.8, line = list(width = 2, color = 'black')),
    color = ~Year,  # Optional: Use a color scale by year to visually distinguish them
    hoverinfo = "text",
    text = ~paste(
      "<b>Year:</b>", Year,
      "<br><b>Total U.S. GDP:</b> $", round(Total_GDP, 1), " billion",
      "<br><b>Transportation GDP:</b> $", round(Transportation_GDP, 1), " billion",
      "<br><b>Air:</b> $", round(Air, 1), " billion",
      "<br><b>Rail:</b> $", round(Rail, 1), " billion",
      "<br><b>Water:</b> $", round(Water, 1), " billion",
      "<br><b>Truck:</b> $", round(Truck, 1), " billion",
      "<br><b>Transit:</b> $", round(Transit, 1), " billion",
      "<br><b>Pipeline:</b> $", round(Pipeline, 1), " billion",
      "<br><b>Support Activities:</b> $", round(Support, 1), " billion",
      "<br><b>Warehousing:</b> $", round(Warehousing, 1), " billion",
      "<br><b>Transportation as % of Total GDP:</b> ", round(Percent_Transport_GDP, 1), "%"
    )
  ) %>%
  # External factors layer
  add_trace(
    data = external_factors,
    x = ~Year,
    y = ~0.2,  # position external events slightly above GDP points
    type = "scatter",
    mode = "markers+text",
    marker = list(size = 10, color = "red"),
    text = ~Event,
    textposition = "top center",
    hoverinfo = "text",
    textfont = list(size = 10),
    hovertext = ~paste(
      "<b>Event:</b>", Event,
      "<br><b>Description:</b>", Description,
      "<br><b>Category:</b>", Category
    )
  ) %>%
  layout(
    title = list(
      text = "<b>Interactive Timeline of GDP Contributions and Major External Factors Affecting the U.S. Transportation Sector</b>",
      font = list(size = 18)
    ),
    xaxis = list(
      title = "Year",
      tickmode = "linear",
      showgrid = FALSE,
      zeroline = FALSE,
      tickangle = 45,
      dtick = 5
    ),
    yaxis = list(
      title = "",
      showticklabels = FALSE,
      zeroline = FALSE
    ),
    hoverlabel = list(
      bgcolor = "white",
      font = list(size = 12)
    ),
    showlegend = FALSE
  )

# Display the plot
timeline_plot

```

```{r}



# Define colors for each sector and set different y-positions for external events based on index
sector_colors <- c("Air" = "blue", "Rail" = "green", "Water" = "purple", "Truck" = "orange")

# Further increase y position offset for external events
external_factors$y_position <- rep(c(-0.4, -0.2), length.out = nrow(external_factors))  # Increased offset

# Create the interactive plot with further staggered external events
timeline_plot <- plot_ly() %>%
  # Air sector data layer
  add_trace(
    data = timeline_data,
    x = ~Year,
    y = ~0.8,  # Position Air sector data slightly above
    type = "scatter",
    mode = "markers",
    marker = list(size = 10, opacity = 0.8, color = sector_colors["Air"]),
    hoverinfo = "text",
    name = "Air",
    text = ~paste(
      "<b>Year:</b>", Year,
      "<br><b>Air GDP:</b> $", round(Air, 1), " billion"
    )
  ) %>%
  # Rail sector data layer
  add_trace(
    data = timeline_data,
    x = ~Year,
    y = ~0.6,  # Position Rail sector data below Air
    type = "scatter",
    mode = "markers",
    marker = list(size = 10, opacity = 0.8, color = sector_colors["Rail"]),
    hoverinfo = "text",
    name = "Rail",
    text = ~paste(
      "<b>Year:</b>", Year,
      "<br><b>Rail GDP:</b> $", round(Rail, 1), " billion"
    )
  ) %>%
  # Water sector data layer
  add_trace(
    data = timeline_data,
    x = ~Year,
    y = ~0.4,  # Position Water sector data below Rail
    type = "scatter",
    mode = "markers",
    marker = list(size = 10, opacity = 0.8, color = sector_colors["Water"]),
    hoverinfo = "text",
    name = "Water",
    text = ~paste(
      "<b>Year:</b>", Year,
      "<br><b>Water GDP:</b> $", round(Water, 1), " billion"
    )
  ) %>%
  # Truck sector data layer
  add_trace(
    data = timeline_data,
    x = ~Year,
    y = ~0.2,  # Position Truck sector data below Water
    type = "scatter",
    mode = "markers",
    marker = list(size = 10, opacity = 0.8, color = sector_colors["Truck"]),
    hoverinfo = "text",
    name = "Truck",
    text = ~paste(
      "<b>Year:</b>", Year,
      "<br><b>Truck GDP:</b> $", round(Truck, 1), " billion"
    )
  ) %>%
  # External events layer with increased vertical spacing
  add_trace(
    data = external_factors,
    x = ~Year,
    y = ~y_position,  # Alternate y positions for external events with further spacing
    type = "scatter",
    mode = "markers+text",
    marker = list(size = 12, color = "red", symbol = "diamond"),
    text = ~Event,
    textposition = ~Label_Position,
    textfont = list(color = ~rep(label_colors, length.out = nrow(external_factors)), size = 10),  # Adjust font size if needed
    hoverinfo = "text",
    name = "External Events",
    hovertext = ~paste(
      "<b>Event:</b>", Event,
      "<br><b>Description:</b>", Description,
      "<br><b>Category:</b>", Category
    )
  ) %>%
  layout(
    title = list(
      text = "<b>Sector-Specific Timeline of GDP Contributions and Major External Factors</b>",
      font = list(size = 18)
    ),
    xaxis = list(
      title = "Year",
      tickmode = "linear",
      range = c(1970, 2025),  # Slightly expand x-axis range for better spacing
      showgrid = FALSE,
      zeroline = FALSE,
      tickangle = 45,  # Angle x-axis labels slightly
      dtick = 5
    ),
    yaxis = list(
      title = "",
      showticklabels = FALSE,
      zeroline = FALSE
    ),
    hoverlabel = list(
      bgcolor = "white",
      font = list(size = 12)
    ),
    showlegend = TRUE
  )

# Display the plot
timeline_plot

```
